---
title: "Getting started with eDNAstructure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with eDNAstructure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Stan-fitting chunks are not evaluated at build time (they compile and
  # sample, which is slow). Set eval = TRUE locally to run them.
  eval = FALSE
)
```

# Overview

`eDNAstructure` fits Bayesian Dirichlet-Multinomial Mixture (DMM) models to eDNA
metabarcoding read counts. Given a sample-by-taxon count matrix and optional
environmental covariates, it identifies latent ecological communities, estimates
their compositions, and quantifies how covariates drive community membership.

This vignette is a short tour. For full argument-by-argument documentation see
the function help pages (e.g. `?eDNA_dmm`) and the package README.

```{r load}
library(eDNAstructure)
```

# Example data

The package ships a small simulated dataset with known ground truth:

```{r data}
data <- get_example_data()
dim(data$counts)        # 20 samples x 40 taxa
head(data$covariates)
```

You can also simulate your own dataset with controllable structure:

```{r simulate}
sim <- simulate_eDNA_survey(
  n_communities         = 4,
  n_species             = 40,
  samples_per_community = 5,
  seed                  = 2026
)
```

# Inspect raw composition before fitting

```{r raw}
plot_true_compositions(
  data$counts,
  metadata  = data$covariates,
  facet_var = "TrueCommunity"
)
```

# Fit the model

```{r fit}
fit <- eDNA_dmm(
  counts     = data$counts,
  covariates = data$covariates[, c("Depth", "Distance_shore")],
  K          = 4
)

print(fit)
summary(fit)
```

The first call compiles the Stan model (or loads the precompiled one shipped
with the package); subsequent calls go straight to sampling.

# Choosing K with LOO cross-validation

```{r loo}
loo_result <- eDNA_loo(
  data$counts,
  data$covariates[, c("Depth", "Distance_shore")],
  K_range = 2:5
)
loo_result$plot

# All fitted models are retained — no need to refit:
fit4 <- loo_result$fits[["K4"]]
```

# Visualize the results

```{r viz}
# STRUCTURE-style assignment bars
eDNA_dmm_structure(fit, metadata = data$covariates,
                   facet_var = "TrueCommunity", sort_var = "Depth")

# NMDS ordination colored by community
eDNA_dmm_nmds(fit)$plot

# Prior vs posterior covariate effects
eDNA_dmm_beta(fit)$plot

# Posterior community compositions
eDNA_dmm_compositions(fit)
```

# Where to go next

- `?eDNA_dmm` — every fitting argument explained
- `?eDNA_dmm_structure`, `?eDNA_dmm_nmds`, `?eDNA_dmm_beta` — visualization options
- `?simulate_eDNA_survey` — the full simulation pipeline
